MLExplain

### 2018-05-28 paper.tex
Continué de rédigé le papier
Ajout d'une fichier recensant les lignes de code, commentaires et appels à bind
Ajouté un tableaux de statistiques sur le projet
first draft

%\documentclass{article}
\documentclass[twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,shapes,arrows.meta}

@@

The core of JSExplain is a JavaScript interpreter written in a subset of OCaml.
This interpreter is compiled to a subset of JavaScript and instrumented to
generate a trace of its execution when run. This trace can then be navigated
using a web-based
tool.\footnote{https://jscert.github.io/jsexplain/branch/master/driver.html}

@@

At first glance, JSExplain seems tightly linked to JavaScript. However, the
architecture of the project actually allows us to take off the interpreter
completely and replace it with another one.

% Plus de blabla

@@

To interpret OCaml code, we need to parse and type it. However, our compiler
does not allow us to use the standard library nor an external one. Fortunately,
JSExplain itself did not trace both parsing and typing processes. That is why
we have been able to separate totally the frontend from the interpreter itself
and use another compiler to compile it. The figure \ref{fig:archi} describes
the architecture of the whole application.

The frontend is compiled with \emph{js\_of\_ocaml}
\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
\emph{Ocsigen} \cite{balat:hal-00691710}. The largest part of the frontend is a
module that serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a
JavaScript object compatible with the backend interpreter's own representation
of the AST. We use the \texttt{compiler-libs} library to do the actual parsing
and typing. We ensure that the typed AST we get is the same as the official
OCaml compiler AST since we use the exact same frontend.

\subsection{Solutions to compiler limitations}

As stated before, the backend of our interpreter had to be written in
\texttt{fOCaml} to be integrated into the web interface. This limitation forced
to write non-idiomatic OCaml.

% Toute une partie sur MLExplain lui-même...

@@

The data structures manipulated by MLExplain are very different from those
manipulated by JSExplain since they don't interpret the same language. This led
us to modify the web interface so that it can recognize the different syntactic
elements of OCaml and pretty print them correctly as well as OCaml values.

Fortunately the web interface is very simple. It is composed of some auxiliary
JavaScript files, the main JavaScript file -- \texttt{navig-driver.js} --
containing the logic of the application and the corresponding HTML page. The
only modified file is \texttt{navig-driver.js}.

>>>>

%\documentclass{article}
\documentclass[twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,shapes,arrows.meta}
\usepackage{hyperref}

@@

The core of JSExplain is a JavaScript interpreter written in a subset of OCaml.
This interpreter is compiled to a subset of JavaScript and instrumented to
generate a trace of its execution when run. This trace can then be navigated
using a web-based
tool.\footnote{\url{https://jscert.github.io/jsexplain/branch/master/driver.html}}

@@

At first glance, JSExplain seems tightly linked to JavaScript. However, the
architecture of the project is quite modular, and can be used for a different
programming language, by providing an interpreter written in the subset of OCaml
that we support as well as a way to display OCaml values.

@@

To interpret OCaml code, we first need to parse and type it. However, our
compiler does not allow us to use the standard library nor an external one. As
we do not need to trace the parsing nor the typing, we have been able to
separate totally the frontend from the interpreter itself and use another
compiler to compile it. Figure \ref{fig:archi} describes the architecture of
the whole application.

The frontend is compiled with \emph{js\_of\_ocaml}
\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project \emph{Ocsigen}
\cite{balat:hal-00691710}. The largest part of the frontend is a module that
serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a JavaScript
object compatible with the backend interpreter's own representation of the AST.
We use the OCaml compiler, namely the \texttt{compiler-libs} library, to do the
actual parsing and typing. We thus have the guarantee that the typed AST we get
is the same as the official OCaml compiler AST.

Unfortunately, we cannot directly use the typed AST, as \emph{js\_of\_ocaml}
takes OCaml bytecode as input, which does not include sufficient information
such as the names of constructors. After parsing and typing, we thus navigate
the AST and call JavaScript functions that create the data in the expected format.

\subsection{Compiler limitations}

As stated before, the backend of our interpreter is written in a purely
functional strict subset of OCaml. As a consequence, it was necessary to encode
some features, such as exceptions. To this end, we use a monadic approach and
rely on \texttt{ppx} extensions to simplify the syntax.

As an example, consider the \texttt{option} monad, whose main operator is
defined as follows.

\begin{verbatim}
let bind_option opt f = match opt with
| Some value -> f value
| None -> None
\end{verbatim}

To use this operator, one would need to write the following.
\begin{verbatim}
bind_option
  (function_that_could_fail 0)
  (fun a -> continuation a)
\end{verbatim}
Such code can become difficult to read, especially when one is chaining several
functions.

The use of \texttt{ppx} lets us write the much simpler code, which is
transformed to the previous version during parsing.
\begin{verbatim}
let%some a = function_that_could_fail 0 in
continuation a
\end{verbatim}

@@

The data structures manipulated by MLExplain are very different from those
manipulated by JSExplain since they don't interpret the same language. We thus
had to modify the web interface so that it can recognize the different syntactic
elements of OCaml and pretty print them correctly as well as OCaml values.

Fortunately the web interface is very simple. It is composed of some auxiliary
JavaScript files, the main JavaScript file -- \texttt{navig-driver.js} --
containing the logic of the application and the corresponding HTML page. The
only modified file is \texttt{navig-driver.js}. The resulting tool can be tested
online.\footnote{\url{https://docteur-lalla.github.io/mlexplain/branch/master/driver.html}}

\section{Statistics}

We give some statistics about the project in Figure~\ref{fig:statistics}. As can
be seen, the largest part of the code is the frontend, in particular the
conversion from OCaml's AST to our format. The interpreter is quite short,
because the execution of OCaml is actually very simple. Note that we support all
of OCaml, with the exception of objects.

\begin{figure}[hb]
  \begin{tabular}{| c | c | c | c |}
    \hline
    & Comments & Code & Bind calls \\
    \hline
    Interpreter & 164 & 394 & 62 \\
    Primitives & 31 & 35 & 0 \\
    Execution context & 22 & 13 & 0 \\
    Frontend & 41 & 542 & 0 \\
    Syntax definition & 17 & 65 & 0 \\
    Auxiliary files & 72 & 232 & 2 \\
    \hline
    Total (no .mli) & 392 & 1621 & 70 \\
    Total (with .mli) & 401 & 1636 & 70 \\
    \hline
  \end{tabular}

  \centering
  \caption{Statistics about the project's code}
  \label{fig:statistics}
\end{figure}

\section{Conclusion}
We have shown how JSExplain can be easily adapted to other programming language,
in this case OCaml. Most of the complexity of OCaml being its typing, writing
the interpreter was quite straightforward. As future work, we plan to extend the
subset of OCaml supported by our compiler, to be able to directly use the
standard library when writing interpreters.
